---
title: "03 - SARIMA Models (Manual Analysis)"
author: "Time Series Analysis"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Introduction

This notebook implements SARIMA (Seasonal ARIMA) models with proper ACF/PACF analysis to determine the optimal model order, following Box-Jenkins methodology.

# Load Required Libraries

```{r libraries}
if (!require("forecast")) install.packages("forecast")
if (!require("tseries")) install.packages("tseries")
if (!require("ggplot2")) install.packages("ggplot2")
if (!require("Metrics")) install.packages("Metrics")

library(forecast)
library(tseries)
library(ggplot2)
library(Metrics)
```

# Load Data

```{r load-data}
load("../data/elec_data.RData")

cat("Training set length:", length(train_ts), "\n")
cat("Validation set length:", length(validation_ts), "\n")
cat("Forecast horizon:", horizon, "\n")
cat("Seasonal period:", frequency_15min, "\n")
```

# Time Series Analysis

## Visual Inspection

```{r visual-inspection, fig.width=12, fig.height=4}
autoplot(train_ts) +
  ggtitle("Training Data - Electricity Consumption") +
  xlab("Time") + ylab("Power (kW)")
```

## Initial ACF/PACF Analysis

```{r initial-acf, fig.width=12, fig.height=8}
# Display ACF and PACF
ggtsdisplay(train_ts, lag.max = 200, main = "Original Series - ACF/PACF")
```

**Observations:**
- Clear seasonality at lag 96 (daily pattern)
- The ACF shows slow decay → non-stationarity
- Need to apply seasonal differencing

# Differencing Analysis

## Seasonal Differencing (D=1)

```{r seasonal-diff, fig.width=12, fig.height=8}
# Apply seasonal differencing (lag = 96)
train_sdiff <- diff(train_ts, lag = frequency_15min)

ggtsdisplay(train_sdiff, lag.max = 200, main = "After Seasonal Differencing (D=1)")
```

We eliminated seasonality. Let's check if we need additional differencing.

```{r double-diff, fig.width=12, fig.height=8}
# Apply first differencing to check
train_ddiff <- diff(train_sdiff)

ggtsdisplay(train_ddiff, lag.max = 100, main = "After Seasonal + First Differencing (D=1, d=1)")
```

```{r ljung-box}
# Ljung-Box test for autocorrelation
Box.test(train_sdiff, lag = 20, type = "Ljung-Box")
```

# Model Building Based on ACF/PACF

Based on the ACF/PACF analysis:

- **ACF**: Shows significant spike at lag 96 (seasonal MA component) with gradual decay
- **PACF**: Shows significant spikes at early lags (AR component)

We'll start with a simple model and iteratively improve:

## Model 1: SARIMA(0,0,0)(0,1,1)[96]

```{r model1, fig.width=12, fig.height=6}
sarima_011 <- Arima(train_ts, order = c(0, 0, 0), 
                    seasonal = list(order = c(0, 1, 1), period = frequency_15min))

cat("Model 1: SARIMA(0,0,0)(0,1,1)[96]\n")
cat("AIC:", sarima_011$aic, "\n")

# Check residuals
sarima_011 %>% residuals() %>% ggtsdisplay(lag.max = 100, 
                                            main = "Residuals - SARIMA(0,0,0)(0,1,1)")
```

Still correlation in residuals. Looking at PACF, we see significant spikes around lag 5.

## Model 2: SARIMA(5,0,0)(0,1,1)[96]

```{r model2, fig.width=12, fig.height=6}
sarima_501 <- Arima(train_ts, order = c(5, 0, 0), 
                    seasonal = list(order = c(0, 1, 1), period = frequency_15min))

cat("\nModel 2: SARIMA(5,0,0)(0,1,1)[96]\n")
cat("AIC:", sarima_501$aic, "\n")

# Check residuals
sarima_501 %>% residuals() %>% ggtsdisplay(lag.max = 100, 
                                            main = "Residuals - SARIMA(5,0,0)(0,1,1)")
```

```{r check-residuals-2}
checkresiduals(sarima_501)
```

## Model 3: SARIMA(5,0,1)(0,1,1)[96]

```{r model3, fig.width=12, fig.height=6}
sarima_511 <- Arima(train_ts, order = c(5, 0, 1), 
                    seasonal = list(order = c(0, 1, 1), period = frequency_15min))

cat("\nModel 3: SARIMA(5,0,1)(0,1,1)[96]\n")
cat("AIC:", sarima_511$aic, "\n")

# Check residuals
checkresiduals(sarima_511)
```

## Auto ARIMA for Comparison

```{r auto-arima}
cat("\nRunning auto.arima for comparison...\n")
sarima_auto <- auto.arima(train_ts, seasonal = TRUE, stepwise = TRUE, approximation = TRUE)

cat("\nAuto ARIMA selected:", capture.output(sarima_auto)[2], "\n")
cat("AIC:", sarima_auto$aic, "\n")

checkresiduals(sarima_auto)
```

# Model Comparison

```{r model-summary}
# Summary of all models
cat("\n=== MODEL COMPARISON ===\n\n")
cat("SARIMA(0,0,0)(0,1,1)[96] - AIC:", sarima_011$aic, "\n")
cat("SARIMA(5,0,0)(0,1,1)[96] - AIC:", sarima_501$aic, "\n")
cat("SARIMA(5,0,1)(0,1,1)[96] - AIC:", sarima_511$aic, "\n")
cat("Auto ARIMA              - AIC:", sarima_auto$aic, "\n")
```

# Forecasting

```{r forecasts, fig.width=12, fig.height=8}
# Generate forecasts
fc_011 <- forecast(sarima_011, h = horizon)
fc_501 <- forecast(sarima_501, h = horizon)
fc_511 <- forecast(sarima_511, h = horizon)
fc_auto <- forecast(sarima_auto, h = horizon)

actual <- as.numeric(validation_ts)

# Plot comparison
par(mfrow = c(2, 2))
plot(fc_011, main = "SARIMA(0,0,0)(0,1,1)")
lines(c(rep(NA, length(train_ts)), actual), col = "red")

plot(fc_501, main = "SARIMA(5,0,0)(0,1,1)")
lines(c(rep(NA, length(train_ts)), actual), col = "red")

plot(fc_511, main = "SARIMA(5,0,1)(0,1,1)")
lines(c(rep(NA, length(train_ts)), actual), col = "red")

plot(fc_auto, main = paste("Auto:", capture.output(sarima_auto)[2]))
lines(c(rep(NA, length(train_ts)), actual), col = "red")
par(mfrow = c(1, 1))
```

# Metrics Comparison

```{r metrics}
calc_metrics <- function(actual, predicted, model_name, aic) {
  data.frame(
    Model = model_name,
    RMSE = rmse(actual, as.numeric(predicted)),
    MAE = mae(actual, as.numeric(predicted)),
    MAPE = mape(actual, as.numeric(predicted)) * 100,
    AIC = aic
  )
}

results <- rbind(
  calc_metrics(actual, fc_011$mean, "SARIMA(0,0,0)(0,1,1)", sarima_011$aic),
  calc_metrics(actual, fc_501$mean, "SARIMA(5,0,0)(0,1,1)", sarima_501$aic),
  calc_metrics(actual, fc_511$mean, "SARIMA(5,0,1)(0,1,1)", sarima_511$aic),
  calc_metrics(actual, fc_auto$mean, "Auto SARIMA", sarima_auto$aic)
)

print(results)

best_idx <- which.min(results$RMSE)
cat("\nBest model by RMSE:", results$Model[best_idx], "\n")
```

# Visualization

```{r final-viz, fig.width=12, fig.height=6}
# Zoomed comparison
plot_data <- data.frame(
  Time = 1:horizon,
  Actual = actual,
  `SARIMA_501` = as.numeric(fc_501$mean),
  `SARIMA_511` = as.numeric(fc_511$mean),
  `Auto_SARIMA` = as.numeric(fc_auto$mean),
  check.names = FALSE
)

library(tidyr)
plot_long <- plot_data %>%
  pivot_longer(cols = -Time, names_to = "Model", values_to = "Power")

ggplot(plot_long, aes(x = Time, y = Power, color = Model, linetype = Model)) +
  geom_line(linewidth = 0.8) +
  scale_linetype_manual(values = c("Actual" = "solid", 
                                   "SARIMA_501" = "dashed",
                                   "SARIMA_511" = "dashed",
                                   "Auto_SARIMA" = "dashed")) +
  labs(title = "SARIMA Models Comparison",
       x = "Time (15-min intervals)",
       y = "Power (kW)") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

# Save Results

```{r save-results}
# Select best forecast
best_forecasts <- list(fc_011, fc_501, fc_511, fc_auto)
best_models <- list(sarima_011, sarima_501, sarima_511, sarima_auto)

sarima_results <- list(
  models = list(
    sarima_011 = sarima_011,
    sarima_501 = sarima_501,
    sarima_511 = sarima_511,
    sarima_auto = sarima_auto
  ),
  forecasts = list(
    sarima_011 = as.numeric(fc_011$mean),
    sarima_501 = as.numeric(fc_501$mean),
    sarima_511 = as.numeric(fc_511$mean),
    sarima_auto = as.numeric(fc_auto$mean)
  ),
  metrics = results,
  best_model = results$Model[best_idx],
  best_rmse = results$RMSE[best_idx],
  best_forecast = as.numeric(best_forecasts[[best_idx]]$mean),
  best_model_object = best_models[[best_idx]]
)

save(sarima_results, file = "../data/sarima_results.RData")
cat("Results saved to ../data/sarima_results.RData\n")
```

# Session Info

```{r session-info}
sessionInfo()
```
