---
title: "09 - Final Report: Model Comparison and Prediction"
author: "Time Series Analysis"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Executive Summary

This report compares all time series forecasting models tested for electricity consumption prediction and selects the best model to generate the final forecast for 2/19/2010.

# Load Required Libraries

```{r libraries}
if (!require("ggplot2")) install.packages("ggplot2")
if (!require("dplyr")) install.packages("dplyr")
if (!require("knitr")) install.packages("knitr")
if (!require("kableExtra")) install.packages("kableExtra")
if (!require("writexl")) install.packages("writexl")
if (!require("R6")) install.packages("R6")
if (!require("tidyr")) install.packages("tidyr")
if (!require("xgboost")) install.packages("xgboost")
if (!require("e1071")) install.packages("e1071")
if (!require("forecast")) install.packages("forecast")

library(ggplot2)
library(dplyr)
library(knitr)
library(kableExtra)
library(writexl)
library(R6)
library(tidyr)
library(xgboost)
library(e1071)
library(forecast)

# Load WNN
tryCatch({
  library(OlivierBorot)
}, error = function(e) {
  source("../../R/WNN.R")
})
```

# Load All Results

```{r load-results}
# Load original data
load("../data/elec_data.RData")

# Load all model results
results_files <- c(
  "hw_results.RData",
  "arima_results.RData", 
  "sarima_results.RData",
  "auto_sarima_results.RData",
  "wnn_results.RData",
  "rnn_results.RData",
  "svm_results.RData",
  "xgb_results.RData"
)

loaded_results <- list()

for (f in results_files) {
  file_path <- paste0("../data/", f)
  if (file.exists(file_path)) {
    load(file_path)
    model_name <- gsub("_results.RData", "", f)
    cat("Loaded:", f, "\n")
  } else {
    cat("Missing:", f, "- Please run the corresponding notebook first.\n")
  }
}
```

# Model Comparison

## Collect All Metrics

```{r collect-metrics}
# Collect best results from each model family
all_results <- data.frame()

# Holt-Winters
if (exists("hw_results")) {
  all_results <- rbind(all_results, data.frame(
    Category = "Exponential Smoothing",
    Model = hw_results$best_model,
    RMSE = hw_results$best_rmse,
    stringsAsFactors = FALSE
  ))
}

# ARIMA
if (exists("arima_results")) {
  all_results <- rbind(all_results, data.frame(
    Category = "ARIMA",
    Model = arima_results$best_model,
    RMSE = arima_results$best_rmse,
    stringsAsFactors = FALSE
  ))
}

# SARIMA
if (exists("sarima_results")) {
  all_results <- rbind(all_results, data.frame(
    Category = "SARIMA",
    Model = sarima_results$best_model,
    RMSE = sarima_results$best_rmse,
    stringsAsFactors = FALSE
  ))
}

# Auto SARIMA
if (exists("auto_sarima_results")) {
  all_results <- rbind(all_results, data.frame(
    Category = "Auto SARIMA",
    Model = auto_sarima_results$best_model,
    RMSE = auto_sarima_results$best_rmse,
    stringsAsFactors = FALSE
  ))
}

# WNN
if (exists("wnn_results")) {
  all_results <- rbind(all_results, data.frame(
    Category = "WNN",
    Model = wnn_results$best_model_name,
    RMSE = wnn_results$best_rmse,
    stringsAsFactors = FALSE
  ))
}

# RNN/LSTM
if (exists("rnn_results_save")) {
  all_results <- rbind(all_results, data.frame(
    Category = "RNN/LSTM",
    Model = rnn_results_save$best_model_name,
    RMSE = rnn_results_save$best_rmse,
    stringsAsFactors = FALSE
  ))
}

# SVM
if (exists("svm_results_save")) {
  all_results <- rbind(all_results, data.frame(
    Category = "SVM",
    Model = svm_results_save$best_model_name,
    RMSE = svm_results_save$best_rmse,
    stringsAsFactors = FALSE
  ))
}

# XGBoost
if (exists("xgb_results_save")) {
  all_results <- rbind(all_results, data.frame(
    Category = "XGBoost",
    Model = xgb_results_save$best_model_name,
    RMSE = xgb_results_save$best_rmse,
    stringsAsFactors = FALSE
  ))
}

# Sort by RMSE
all_results <- all_results[order(all_results$RMSE), ]
all_results$Rank <- 1:nrow(all_results)

# Display table
kable(all_results[, c("Rank", "Category", "Model", "RMSE")], 
      caption = "All Models Comparison - Ranked by RMSE",
      digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

## Best Model Selection

```{r best-model}
best_category <- all_results$Category[1]
best_model_name <- all_results$Model[1]
best_rmse <- all_results$RMSE[1]

cat("==============================================\n")
cat("           BEST MODEL SELECTED\n")
cat("==============================================\n")
cat("Category:", best_category, "\n")
cat("Model:", best_model_name, "\n")
cat("RMSE:", round(best_rmse, 2), "\n")
cat("==============================================\n")
```

## Visualization: Model Comparison

```{r comparison-plot, fig.width=10, fig.height=6}
ggplot(all_results, aes(x = reorder(Model, RMSE), y = RMSE, fill = Category)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = round(RMSE, 1)), hjust = -0.1, size = 3) +
  coord_flip() +
  labs(title = "Model Comparison by RMSE",
       x = "Model",
       y = "RMSE (kW)",
       fill = "Category") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Forecast Comparison on Validation Set

```{r forecast-comparison, fig.width=12, fig.height=8}
actual <- as.numeric(validation_ts)

# Collect all best forecasts
forecast_data <- data.frame(
  Time = 1:horizon,
  Actual = actual
)

if (exists("hw_results")) forecast_data$HW <- hw_results$best_forecast
if (exists("arima_results")) forecast_data$ARIMA <- arima_results$best_forecast
if (exists("sarima_results")) forecast_data$SARIMA <- sarima_results$best_forecast
if (exists("auto_sarima_results")) forecast_data$Auto_SARIMA <- auto_sarima_results$best_forecast
if (exists("wnn_results")) forecast_data$WNN <- wnn_results$best_forecast
if (exists("rnn_results_save")) forecast_data$RNN <- rnn_results_save$best_forecast
if (exists("svm_results_save")) forecast_data$SVM <- svm_results_save$best_forecast
if (exists("xgb_results_save")) forecast_data$XGBoost <- xgb_results_save$best_forecast

# Reshape for plotting
forecast_long <- forecast_data %>%
  pivot_longer(cols = -Time, names_to = "Model", values_to = "Value")

ggplot(forecast_long, aes(x = Time, y = Value, color = Model, linetype = Model)) +
  geom_line(linewidth = 0.8) +
  scale_linetype_manual(values = c("Actual" = "solid", 
                                   "HW" = "dashed", 
                                   "ARIMA" = "dashed",
                                   "SARIMA" = "dashed",
                                   "Auto_SARIMA" = "dashed",
                                   "WNN" = "dashed",
                                   "RNN" = "dashed",
                                   "SVM" = "dashed",
                                   "XGBoost" = "dashed")) +
  scale_color_manual(values = c("Actual" = "black",
                                "HW" = "blue",
                                "ARIMA" = "red",
                                "SARIMA" = "green",
                                "Auto_SARIMA" = "purple",
                                "WNN" = "orange",
                                "RNN" = "cyan",
                                "SVM" = "brown",
                                "XGBoost" = "darkgreen")) +
  labs(title = "All Models - Forecast vs Actual (Validation Set)",
       x = "Time (15-min intervals)",
       y = "Power (kW)") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

# Generate Final Forecast for 2/19/2010

The final forecast will be generated using the best performing model.

```{r final-forecast}
# Use the FULL dataset (including validation) to train the final model
# and predict the next 96 values (2/19/2010)

cat("Generating final forecast using:", best_category, "-", best_model_name, "\n\n")

# Train on full data (power_ts) and predict next day
if (best_category == "Exponential Smoothing") {
  library(forecast)
  
  if (grepl("Additive", best_model_name)) {
    final_model <- HoltWinters(power_ts, seasonal = "additive")
  } else if (grepl("Multiplicative", best_model_name)) {
    final_model <- HoltWinters(power_ts, seasonal = "multiplicative")
  } else {
    final_model <- ets(power_ts)
  }
  final_forecast <- as.numeric(forecast(final_model, h = horizon)$mean)
  
} else if (best_category == "ARIMA") {
  library(forecast)
  
  # Parse ARIMA order from model name
  order_match <- regmatches(best_model_name, regexec("\\((\\d+),(\\d+),(\\d+)\\)", best_model_name))[[1]]
  p <- as.integer(order_match[2])
  d <- as.integer(order_match[3])
  q <- as.integer(order_match[4])
  
  final_model <- Arima(power_ts, order = c(p, d, q))
  final_forecast <- as.numeric(forecast(final_model, h = horizon)$mean)
  
} else if (best_category == "SARIMA") {
  library(forecast)
  # Refit the best SARIMA model on full data
  final_model <- auto.arima(power_ts, seasonal = TRUE, stepwise = TRUE, approximation = TRUE)
  final_forecast <- as.numeric(forecast(final_model, h = horizon)$mean)
  
} else if (best_category == "Auto SARIMA") {
  library(forecast)
  final_model <- auto.arima(power_ts, stepwise = TRUE, approximation = TRUE)
  final_forecast <- as.numeric(forecast(final_model, h = horizon)$mean)
  
} else if (best_category == "WNN") {
  # Use best WNN parameters
  w <- wnn_results$best_params$window
  k <- wnn_results$best_params$k
  
  final_wnn <- WNN$new(horizon = horizon, window = w, k = k)
  final_forecast <- as.numeric(final_wnn$fit_predict(power_ts))
  
} else if (best_category == "RNN/LSTM") {
  # For RNN, we need to retrain - use the best forecast from validation as approximation
  # In production, you would retrain the neural network on the full dataset
  cat("Note: Using pre-computed RNN forecast. For production, retrain on full data.\n")
  # Approximate by shifting the validation forecast pattern
  final_forecast <- rnn_results_save$best_forecast
  
} else if (best_category == "SVM") {
  # Retrain SVM on full data
  train_data <- as.numeric(power_ts)
  train_mean <- mean(train_data)
  train_sd <- sd(train_data)
  train_scaled <- (train_data - train_mean) / train_sd
  
  lags <- svm_results_save$best_params$lags
  
  # Create lagged features function
  create_lagged <- function(data, lags) {
    n <- length(data)
    max_lag <- max(lags)
    X <- matrix(NA, nrow = n - max_lag, ncol = length(lags))
    for (i in seq_along(lags)) X[, i] <- data[(max_lag - lags[i] + 1):(n - lags[i])]
    y <- data[(max_lag + 1):n]
    list(X = X, y = y, max_lag = max_lag)
  }
  
  lag_data <- create_lagged(train_scaled, lags)
  train_df <- data.frame(y = lag_data$y, lag_data$X)
  
  final_model <- svm(y ~ ., data = train_df, kernel = "radial", cost = 10, gamma = 0.1)
  
  # Recursive forecast
  n <- length(train_scaled)
  predictions <- numeric(horizon)
  extended <- c(train_scaled, rep(NA, horizon))
  for (h in 1:horizon) {
    idx <- n + h
    features <- sapply(lags, function(lag) extended[idx - lag])
    pred <- predict(final_model, newdata = data.frame(t(features)))
    predictions[h] <- pred
    extended[idx] <- pred
  }
  final_forecast <- predictions * train_sd + train_mean
  
} else if (best_category == "XGBoost") {
  # Retrain XGBoost on full data
  train_data <- as.numeric(power_ts)
  lags <- xgb_results_save$best_params$lags
  
  # Create features
  create_features <- function(data, lags) {
    n <- length(data)
    max_lag <- max(lags)
    X <- matrix(NA, nrow = n - max_lag, ncol = length(lags))
    for (i in seq_along(lags)) X[, i] <- data[(max_lag - lags[i] + 1):(n - lags[i])]
    y <- data[(max_lag + 1):n]
    list(X = as.data.frame(X), y = y, max_lag = max_lag)
  }
  
  feat_data <- create_features(train_data, lags)
  dtrain <- xgb.DMatrix(data = as.matrix(feat_data$X), label = feat_data$y)
  
  params <- list(objective = "reg:squarederror", eta = 0.1, max_depth = 6)
  final_model <- xgb.train(params = params, data = dtrain, nrounds = 100, verbose = 0)
  
  # Recursive forecast
  n <- length(train_data)
  predictions <- numeric(horizon)
  extended <- c(train_data, rep(NA, horizon))
  for (h in 1:horizon) {
    idx <- n + h
    features <- sapply(lags, function(lag) extended[idx - lag])
    pred <- predict(final_model, newdata = matrix(features, nrow = 1))
    predictions[h] <- pred
    extended[idx] <- pred
  }
  final_forecast <- predictions
}

cat("Final forecast generated successfully!\n")
cat("Length:", length(final_forecast), "values\n")
cat("First 5 values:", head(final_forecast, 5), "\n")
cat("Last 5 values:", tail(final_forecast, 5), "\n")
```

# Visualize Final Forecast

```{r final-viz, fig.width=12, fig.height=6}
# Create time index for the forecast day
forecast_times <- seq(
  from = as.POSIXct("2010-02-19 00:15:00"),
  by = "15 min",
  length.out = horizon
)

final_df <- data.frame(
  DateTime = forecast_times,
  Power_kW = final_forecast
)

# Plot
ggplot(final_df, aes(x = DateTime, y = Power_kW)) +
  geom_line(color = "steelblue", linewidth = 1) +
  labs(title = "Final Forecast for 2/19/2010",
       subtitle = paste("Model:", best_model_name),
       x = "Time",
       y = "Power (kW)") +
  theme_minimal() +
  scale_x_datetime(date_labels = "%H:%M", date_breaks = "2 hours")
```

# Export Results to Excel

```{r export-excel}
# Create the result dataframe with only the power values
result_df <- data.frame(
  Power_kW = final_forecast
)

# Save to Excel
output_path <- "../data/result.xlsx"
write_xlsx(result_df, output_path)

cat("==============================================\n")
cat("      RESULTS EXPORTED SUCCESSFULLY\n")
cat("==============================================\n")
cat("File:", output_path, "\n")
cat("Rows:", nrow(result_df), "\n")
cat("Columns:", ncol(result_df), "\n")
cat("==============================================\n")
```

# Summary Statistics

```{r summary-stats}
cat("\n--- Final Forecast Statistics ---\n")
cat("Min:", round(min(final_forecast), 2), "kW\n")
cat("Max:", round(max(final_forecast), 2), "kW\n")
cat("Mean:", round(mean(final_forecast), 2), "kW\n")
cat("Median:", round(median(final_forecast), 2), "kW\n")
cat("Std Dev:", round(sd(final_forecast), 2), "kW\n")
```

# Model Performance Summary

```{r final-summary}
# Create final summary table
summary_table <- all_results %>%
  mutate(
    `Performance` = case_when(
      Rank == 1 ~ "★ BEST",
      Rank == 2 ~ "2nd",
      Rank == 3 ~ "3rd",
      TRUE ~ as.character(Rank)
    )
  ) %>%
  select(Performance, Category, Model, RMSE)

kable(summary_table,
      caption = "Final Model Rankings",
      digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE) %>%
  row_spec(1, bold = TRUE, background = "#d4edda")
```

# Conclusions

Based on our comprehensive comparison of time series forecasting models:

1. **Best Model**: `r best_model_name` from the `r best_category` family
2. **Validation RMSE**: `r round(best_rmse, 2)` kW
3. **Forecast**: 96 values for 2/19/2010 have been exported to `data/result.xlsx`


# Session Info

```{r session-info}
sessionInfo()
```
